---
title: "Stuti_C_HW_NIIN_540"
author: "Stuti C"
date: "2023-09-20"
output: html_document
---

# Part_1: Random variables, normal distributions, variance, covariance, and correlations

Vector of 10 points randomly chosen from the normal distribution with expected mean 10 and standard deviation 5. Calculate the mean and variance.

```{r codeblock SC1, include=TRUE}
v1_sc=rnorm(10, mean = 10, sd = 5)
```

```{r codeblock SC2, include=TRUE}
result.mean <- mean(v1_sc)
print(result.mean)
```
```{r codeblock SC3, include=TRUE}
var=var(v1_sc)
print(var)
```

Plot a histogram of all the points
```{r codeblock SC4, include=TRUE}
v2=rnorm(10, mean = result.mean, sd = var)
h <- hist(v2, breaks=5, plot=TRUE)
plot(h)
```
# repeat with vectors of 20, 50 and 100 points

```{r codeblock SC5, include = TRUE}
v2_sc=rnorm(20, mean = 10, sd = 5)
result.mean <- mean(v2_sc)
print(result.mean)
```

```{r codeblock SC6, include = TRUE}
var=var(v2_sc)
print(var)
```

```{r codeblock SC7, include = TRUE}
v3_sc=rnorm(20, mean = result.mean, sd = var)
h <- hist(v3_sc, breaks=5, plot=TRUE)
plot(h)
```

```{r codeblock SC8, include = TRUE}
v3_sc=rnorm(50, mean = 10, sd = 5)
result.mean <- mean(v3_sc)
print(result.mean)
```

```{r codeblock SC9, include = TRUE}
var=var(v3_sc)
print(var)
```


```{r codeblock SC10, include = TRUE}
v4_sc=rnorm(50, mean = result.mean, sd = var)
h <- hist(v4_sc, breaks=5, plot=TRUE)
plot(h)
```
```{r codeblock SC11, include = TRUE}
v5_sc=rnorm(100, mean = 10, sd = 5)
result.mean <- mean(v5_sc)
print(result.mean)
```
```{r codeblock SC12, include = TRUE}
var=var(v5_sc)
print(var)
```
```{r codeblock SC13, include = TRUE}
v6_sc=rnorm(50, mean = result.mean, sd = var)
h <- hist(v6_sc, breaks=10, plot=TRUE)
plot(h)
```

Create vectors A1 and A2.
A1 is a vector of 150 points (size), with values between range 50-1000. A2 is a vector of 150 points (size) with range 100-1000.
```{r codeblock SC14, include = TRUE}
A1<-runif(150, min = 50, max = 1000)
print(A1)

var=var(A1)
print(var)
```

```{r codeblock SC15, include = TRUE}
A2<-runif(n=150, min = 100, max=1000)
print(A2)

var=var(A2)
print(var)
```

create a 2X2 covariance matrix between A1 and A2. This can be done using the function "cov". First we need to define our dataframe for creating the covariance matrix

```{r codeblock SC16, include = TRUE}
sample_data_vector<-data.frame(var1 = A1, var2 = A2)
cov(sample_data_vector)
```

Covariance is a statistical measure. First, I defined 2 vectors A1 and A2 having the length 150, and then used the above code to create a 2X2 covariance matrix between A1 and A2. Covariance shows the relationship between a pair of random values, and depicts how one variable causes change in the other (measure of linear correlation between the two variables).

Covariance matrix helps us understand how different variables in our dataset relate to each other.

Source:
How to create a covariance matrix <https://www.geeksforgeeks.org/how-to-create-a-covariance-matrix-in-r/>

How to create a vector of numbers: <https://www.statology.org/random-vector-in-r/>

# Calculating correlation matrix between two vectors A1 and A2 using 'Pearson correlation'.

Source: <https://www.geeksforgeeks.org/compute-the-correlation-coefficient-value-between-two-vectors-in-r-programming-cor-function/>

Pearson's correlation gives a coefficient measurement of linear correlation which shows the strength and direction of correlation between variables (positive = similar/convergent, negative = divergent, different). Correlation coefficient value closer to 1 indicates strong correlation.

```{r codeblock SC17, include = TRUE}
x<-c(A1)
y<-c(A2)
print(cor(x, y, method = "pearson"))
```

New vector A3  where each element is the average of that element in vectors A1 and A2.

```{r codeblock SC18, include = TRUE}
x<-c(A1)
y<-c(A2)
A3<- (x+y)/2
print(A3)

var=var(A3)
print(var)
```
```{r codeblock SC19, include = TRUE}
#Check that for an index i, where i is between 1 and 150, A3[i] equals 0.5*(A1[i] + A2[i])
test=0.5*(as.numeric(A1[1:150])+as.numeric(A2[1:150]))
test
```
Changes in variances: the varaince reduces into half since the values of the matrix are also halved.




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
